### 2024-06-05 preregistration/README.qmd
intermediate commit of preregistration

---
title: ""
format: 
  html:
    template: ../assets/template.html
    css: ../assets/style.css
bibliography: ../manuscript/references.bib
csl: ../manuscript/apa.csl
link-citations: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(lmerTest)
library(flextable)
library(ftExtra)

@@

@@

-   Stefan Vermeent^1,2^
-   Anna-Lena Schubert^3^
-   Jean-Louis van Gelder^2^
-   Willem E. Frankenhuis^2,4^

@@

### Q3: Research Questions DONE

**R1:** How is adversity associated with general speed of processing across all cognitive tasks?

**R2:** How is adversity associated with inhibition and attention shifting abilities after accounting for general speed of processing?

**R3:** How is adversity associated with task-general and task-specific response caution?

### Q4: Hypotheses DONE

@@

A1. Adversity is negatively associated with general speed of processing.

A2. Adversity is negatively associated with either or both latent estimates of inhibition and attention shifting, after accounting for general speed of processing.

A3. If latent estimates of inhibition and attention shifting are highly correlated, adversity is negatively associated with a general EF factor spanning both.

@@

B1. Adversity is positively associated with the latent estimate of attention shifting *regardless of the association with inhibition*, after accounting for general speed of processing. A positive association with inhibition would be unexpected yet interesting, and would require revising existing adaptation theoretical frameworks.

B2. The association between adversity and the latent estimate of attention shifting is practically equivalent to zero *and* the association between adversity and the latent estimate of inhibition is negative.

@@

C1. Adversity is associated with more task-general response caution.

@@

### Q5: Dataset DONE

Our study will include 1000 participants who were randomly sampled from the LISS panel (Scherpenzeel, 2011). The LISS panel is a representative probability sample of roughly 5,000 Dutch households (\~7,500 individuals) drawn from the population register by Statistics Netherlands on an invite-only basis. Households without a computer or internet connection are provided with these facilities by LISS. Each year, participants complete the same core battery of questionnaires about---among other topics---their financial situation in the past year. In addition, participants can complete additional online questionnaires every month, with variable content. The current study integrates two data sources. First, our sample of 1000 participants will participate in a new LISS study between May 2024 and July 2024 (hereafter referred to as 'newly collected data'), in which we will include a measure of neighborhood threat, three cognitive tasks measuring attention shifting, two cognitive tasks measuring inhibition, and one cognitive task measuring processing speed. Second, we will access data that were previously collected in LISS (hereafter referred to as 'the LISS archive').

### Q6: Public Availability DONE

All datasets are available in the LISS data archive. Researchers who want to access the data are required to sign a statement confirming that information about individual persons, households, etc., will not be released to others (go to <https://statements.centerdata.nl> for more information).

### Q7: Data Access DONE

@@

After the new data have been collected, they will become available in the general overview at <https://www.dataarchive.lissdata.nl/study-units/view/1>

### Q8: Date of Download DONE

@@

### Q9: Data Collection DONE

General information about recruitment of participants into the LISS study can be found at <https://www.lissdata.nl/methodology>.

Our study will include 1,000 participants who were randomly sampled from the LISS panel. The requirements will be that they are between 20 and 55 years old and have completed at least 1 wave of the Conventional and Computer Crime Victimization study (see Q10).

### Q10: Codebooks DONE

@@

Codebooks for the newly collected data will become available later at <https://github.com/StefanVermeent/liss_ef_2024>

@@

### Q11: Manipulated Variables: DONE

@@

We identified potential confounders of the empirical estimands (i.e., the association between self-reported threat and deprivation with cognitive outcomes) using a Directed Acyclic Graph (DAG).
We analyzed the following variables both conceptually (i.e., based on previous literature and/or our own assumptions) and, in cases of doubt, statistically: (1) age, (2) sex, (3) past adversity exposure (material deprivation and threat combined), (4) recent threat exposure, and (5) recent material deprivation.
Statistical analysis of potential confounders was based on data from a previous LISS study (see Q17 and Q18) using the `dagitty` R package [@textor_2016].
Specifically, we used the `dagitty::localTests()` function to establish whether DAG-implied conditional independencies between certain key variables were supported by the data [@ankan_2021].
The outcome in these conceptual DAGs was working memory capacity, measured on a latent level in the previous LISS study.
Code for these analyses can be found under `scripts/00_dag`.

Figure 1A shows the initially conceptualized DAG. 
Black lines depict our estimands, and red lines depict potential confounding paths.
Solid lines depict paths that we were reasonably sure about given previous literature.
Dashed lines depict paths that we were less sure about, and therefore decided to statistically test as outlined above.

(1) **Age.** *Conceptual reasoning*: Age has a causal effect on processing speed [@salthouse_2016; @salthouse_2019] and on response caution [@starns_2010]. Older adults may on average be more financially stable.

We know of no evidence to consider age as a precedent for threat exposure, especially because we measure recent adversity and not cumulative lifetime adversity (which is likely related to age as older individuals on average will have more cumulative exposure to adverse events).

(2) **Sex.**

(4) **Childhood adversity.** *Conceptual reasoning*:

childhood adversity on education: @fry_2018
Education on poverty: @hofmarcher_2021

```{r}
#| label: Figure1
#| fig-width: 6.25
#| fig-height: 7
#| echo: false
#| dpi: 600
#| out-width: 6in
#| fig-cap: |
#|   **Figure 1.** Directed Acyclic Graph (DAG) depicting the empirical estimands (black lines) and presumed confounding effects (red lines). Age, education level, and childhood adversity are all conceptualized as confounding variables, as they are expected to 
#|   Importantly, we consider deprivation to be a confounder in the effect of threat on cognitive ability, and conversely, we consider threat to be a mediator of the effect of deprivation on cognitive ability [@ning_2023]. From this, it follows that deprivation should be adjusted for when estimating the effect of threat on cognitive ability. However, threat should *not* be adjusted for when estimating the effect of deprivation on cognitive ability.

@@

```{r}
#| echo: false
table1 <- 
  tibble(
    `(Potential) confounder` = c(
      "Age",
      "Education",
      "Sex",
      "Childhood adversity",
      "Recent material deprivation",
      "Recent threat"
    ),
    `Causal effect on cognitive performance` = c(
      "@salthouse_2016; \n @salthouse_2019; \n @starns_2010",
      "@lovden_2020",
      "@ning_2023", 
      "@bos_2009; \n @goodman_2019",
      "*Estimand of interest*",
      "*Estimand of interest*"
    ),
    `Causal effect on recent threat` = c(
      "Unclear",
      "Unclear",
      "@ning_2023", 
      "@bywaters_2016",
      "@bywaters_2016; @ning_2023; ",
      ""
    ),
    `Causal effect on recent material deprivation` = c(
      "Older adults may have more financial resources",
      "@fry_2018",
      "@ning_2023", 
      "@bywaters_2016",
      "",
      ""
    ),
    `Statistical (non-)support` = c(
      paste0("Conditional on childhood adversity and recent deprivation, \n *r* = ", localTest_dag03 |> pull(estimate), " SE = ", localTest_dag03 |> pull(std.error), ", 95% CI = ", localTest_dag03 |> pull(ci), "."),

      paste0("Conditional on childhood adversity, recent deprivation, and age, \n *r* = ", localTest_dag01 |> pull(estimate), " SE = ", localTest_dag01 |> pull(std.error), ", 95% CI = ", localTest_dag01 |> pull(ci), "."),
      "", 
      "",
      paste0("conditional on childhood adversity and age, deprivation is a significant confounder of threat \n *r* = ", localTest_dag04 |> pull(estimate), " SE = ", localTest_dag04 |> pull(std.error), ", 95% CI = ", localTest_dag04 |> pull(ci), "."),
      ""
    ),
    `Confounder in threat model` = c(
      "Yes",
      "No",
      "Yes", 
      "Yes",
      "Yes",
      ""
    ),
    `Confounder in deprivation model` = c(
      "Yes",
      "Yes",
      "Yes", 
      "Yes",
      "",
      "No, is a mediator"
    )
  ) |> 
  flextable() |> 
  ftExtra::colformat_md()

@@

@@

@@

We will first fit the model as specified in Figure 2A (see `mod_v1a` in the code block below). Then, we will fit a second version of this model in which a single common EF factor is estimated, covering both inhibition and attention shifting tasks. 
Using `lavaan::anova`, we will then test whether the second, more constrained model leads to worse model fit, qualified as a significant chi squared change test, and an AIC value > 10. 
If so, we will include submodel 2 in our final model.

@@

A power analysis based on @Kretzschmar_2019 indicated that with an alpha of α = 0.05, we will have \> 90% power to detect small effect sizes (β = 0.1) with a sample size ranging between N = 730 and N = 980 when the reliability of the measures is at least moderate (0.6 -- 0.7). Therefore, we aim for a total sample size of N = 1,000.

@@

<div id="refs"></div>

>>>>

---
title: ""
format: 
  html:
    css: D:/repositories/liss_ef_2024/assets/style.css
    template: ../assets/template.html
bibliography: ../manuscript/references.bib
csl: ../manuscript/apa.csl
link-citations: true
editor: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
library(tidyverse)
library(lmerTest)
library(flextable)
library(ftExtra)
library(tinytable)

@@

```{r}
#| include: false

pagedown::chrome_print(input = "preregistration/README.html", output = "preregistration/preregistration.pdf")
```

[click here to download a pdf version](https://github.com/Schecher1/Minecraft-Server-Creator/blob/master/README.md)

@@

-   Stefan Vermeent^1,2^
-   Anna-Lena Schubert^3^
-   Jean-Louis van Gelder^2^?
-   Willem E. Frankenhuis^2,4^

@@

### Q3: Research Questions

-   **R1:** How is adversity associated with general speed of processing across all cognitive tasks?

-   **R2:** How is adversity associated with inhibition and attention shifting abilities after accounting for general speed of processing?

-   **R3:** How is adversity associated with task-general and task-specific response caution?

### Q4: Hypotheses

@@

-   **A1.** Adversity is negatively associated with general speed of processing.

-   **A2.** Adversity is negatively associated with either or both latent estimates of inhibition and attention shifting, after accounting for general speed of processing.

-   **A3.** If latent estimates of inhibition and attention shifting are highly correlated, adversity is negatively associated with a general EF factor spanning both.

@@

-   **B1.** Adversity is positively associated with the latent estimate of attention shifting *regardless of the association with inhibition*, after accounting for general speed of processing.
A positive association with inhibition would be unexpected yet interesting, and would require revising existing adaptation theoretical frameworks.

-   **B2.** The association between adversity and the latent estimate of attention shifting is practically equivalent to zero *and* the association between adversity and the latent estimate of inhibition is negative.

@@

-   **C1.** Adversity is associated with more task-general response caution.

Go back to [Overview].

@@

### Q5: Dataset

The LISS panel is a representative probability sample of roughly 5,000 Dutch households (\~7,500 individuals) drawn from the population register by Statistics Netherlands on an invite-only basis [@scherpenzeel_2011].
Households without a computer or internet connection are provided with these facilities by LISS.
Each year, participants complete the same core battery of questionnaires about---among other topics---their financial situation in the past year.
In addition, participants can complete additional online questionnaires every month, with variable content.

The current study integrates two data sources.
First, we will sample 1,000 participants in a new LISS study, collected between in May and June 2024 (hereafter referred to as 'newly collected data'), in which we will include measures of neighborhood threat and deprivation, three cognitive tasks measuring attention shifting, two cognitive tasks measuring inhibition, and one cognitive task measuring processing speed.
Second, we will access data that were previously collected in LISS (hereafter referred to as 'the LISS archive').
This preregistration is submitted before completion of (and access to) the newly collected data.

### Q6: Public Availability

All previously collected LISS data are available in the LISS data archive.
Researchers who want to access the data are required to sign a statement confirming that information about individual persons, households, etc., will not be released to others (go to <https://statements.centerdata.nl> for more information).

### Q7: Data Access

@@

After the new data have been collected, they will become available in the general overview at <https://www.dataarchive.lissdata.nl/study-units/view/1>.

### Q8: Date of Download

@@

### Q9: Data Collection TBD (fix inclusion criteria)

General information about recruitment of participants into the LISS study can be found at <https://www.lissdata.nl/methodology>.
The inclusion criteria for the new data collection are that people are between 20 and 55 years old and have completed at least 1 wave of the Conventional and Computer Crime Victimization study (see Q10).

### Q10: Codebooks

@@

Codebooks for the newly collected data will become available later at <https://www.dataarchive.lissdata.nl/study-units/view> and <https://github.com/StefanVermeent/liss_ef_2024>.

Go back to [Overview].

@@

### Q11: Manipulated Variables

@@

We used Directed Acyclic Graphs (DAG) to identify potential confounders of the empirical estimands (i.e., the association between self-reported threat and deprivation with cognitive outcomes).
For more information on DAGs, see @rohrer_2018.

We analyzed the following potential confounders both conceptually (i.e., based on previous literature and/or our own assumptions) and, in cases of doubt, statistically (see below): (1) age [@salthouse_2016; @salthouse_2019; @starns_2010], (2) biological sex [@ning_2023], (3) childhood adversity exposure (material deprivation and threat combined) [@bos_2009; @goodman_2019], and (4) recent material deprivation as a confounder for the effect of threat on cognitive performance [@bywaters_2016; @ning_2023].

**Estimand 1: Material deprivation -\> Cognitive performance.** See Figure 2A for a visualization of the DAG for material deprivation.
All models involving material deprivation as the main predictor will control for these four covariates.
In line with recent work, we consider material deprivation to precede threat exposure [@bywaters_2016; @ning_2023].
For this reason, we do not include threat exposure in models where we are interested in the total effect of material deprivation on cognitive performance.

**Estimand 2: Threat exposure -\> Cognitive performance.** See Figure 2B for a visualization of the DAG for threat exposure.
We considered the same set of confounders as for material deprivation.
However, in the case of threat exposure, we were conceptually less sure about the potential effects of age and education on threat exposure.
We also wanted to gather more empirical support for the assumption that material deprivation should be considered as a confounder of the effect of threat exposure on cognitive performance [@bywaters_2016; @ning_2023].

Table 1 offers an overview of conceptual and statistical evidence for and against each potential confounder.
Statistical analysis of potential confounders was based on data from a previous LISS study (see Q17 and Q18) using the `dagitty` R package [@textor_2016].
Specifically, we used the `dagitty::localTests()` function to establish whether DAG-implied conditional independencies between certain key variables were supported by these data [@ankan_2021].
The outcome in these conceptual DAGs was working memory capacity, measured on a latent level in the previous LISS study.
For each of the three confounders under investigation, we removed the causal pathway between that variable and threat exposure from the DAG, and then tested whether this imposed independence (conditional on the other causal effects in the model) was consistent with the data.
Code for these analyses can be found under `scripts/00_dag`.

```{r}
#| label: Figure1
#| fig-width: 6.25
#| fig-height: 7
#| echo: false
#| dpi: 600
#| out-width: 6in
#| fig-cap: |
#|   **Figure 1.** Directed Acyclic Graph (DAG) depicting the empirical estimands (blue lines) and presumed confounding effects (red lines). Importantly, we consider deprivation to be a confounder in the effect of threat on cognitive ability, and conversely, we consider threat to be a mediator of the effect of deprivation on cognitive ability [@ning_2023]. From this, it follows that deprivation should be adjusted for when estimating the effect of threat on cognitive ability. However, threat should *not* be adjusted for when estimating the effect of deprivation on cognitive ability.

@@

```{r}
#| include: false

table1 <- 
  tibble(
    `(Potential) confounder` = c(
      "Age",
      "Education",
      "Sex",
      "Childhood adversity",
      "Recent material deprivation"
    ),
    `Causal effect on cognitive performance` = c(
      "@salthouse_2016 \n @salthouse_2019 \n @starns_2010",
      "@lovden_2020",
      "@ning_2023", 
      "@bos_2009 \n @goodman_2019",
      "*Estimand of interest*"
    ),
    `Causal effect on recent threat` = c(
      "*Unclear*",
      "*Unclear*",
      "@ning_2023", 
      "@bywaters_2016",
      "@bywaters_2016 \n @ning_2023"
    ),
    `Evidence for conditional independence` = c(
      paste0("Conditional on childhood adversity and recent deprivation, \n *r* = ", localTest_dag03 |> pull(estimate), " SE = ", localTest_dag03 |> pull(std.error), ", 95% CI = ", localTest_dag03 |> pull(ci), "."),

      paste0("Conditional on childhood adversity, recent deprivation, and age, \n *r* = ", localTest_dag01 |> pull(estimate), " SE = ", localTest_dag01 |> pull(std.error), ", 95% CI = ", localTest_dag01 |> pull(ci), "."),
      "", 
      "",
      paste0("conditional on childhood adversity and age, deprivation is a significant confounder of threat \n *r* = ", localTest_dag04 |> pull(estimate), " SE = ", localTest_dag04 |> pull(std.error), ", 95% CI = ", localTest_dag04 |> pull(ci), ".")
    ),
    `Confounder in threat model` = c(
      "Yes",
      "No",
      "Yes", 
      "Yes",
      "Yes"
    ),
  ) |> 
  flextable() |> 
  ftExtra::colformat_md() |> 
  mk_par(
    j = 4,
    value = as_paragraph(
      gg_chunk(value = localTest_plot_list, width = 1, height = 1)
    )
  ) |> 
  add_header_row(
    values = " ",
    colwidths = 5
  ) |> # Add a new header row on top. We can use this new row to add the title
  compose(
    i = 1, j = 1,
    as_paragraph(as_b("Table 1. "), "Conceptual and statistical confounder analysis for the threat exposure model."),
    part = "header"
  ) |>
  add_footer_row(
    values = " ",
    colwidths = 5
  ) %>% 
  add_footer_row(
    values = " ",
    colwidths = 5
  ) %>% 
  compose(
    i = 1, j = 1, 
    as_paragraph(as_i("Note: "), "the second and third columns present conceptual support (based on previous literature and/or our own assumptions) in favor of a causal effect between the confounder and threat and cognitive performance. The embedded plots in column 4 present the outcome of statistical analyses that tested against independence between the confounder and threat exposure, conditional on other variables associated with both [see @ankan_2021 and the main text for more details]."), 
    part = "footer"
  ) |> 
  autofit()
```

```{r}
#| echo: false

@@

We found evidence against conditional independence between age and threat.
We also found evidence against independence between material deprivation and threat exposure, conditional on age and childhood adversity.
Finally, we did not find evidence against conditional independence between education and threat.
Therefore, all models involving recent threat exposure as the main predictor will control for age, sex, childhood adversity, and recent material deprivation.

@@

Go back to [Overview].

@@

We will first fit the model as specified in Figure 2A (see `mod_v1a` in the code block below).
Then, we will fit a second version of this model in which a single common EF factor is estimated, covering both inhibition and attention shifting tasks.
Using `lavaan::anova`, we will then test whether the second, more constrained model leads to worse model fit, qualified as a significant chi squared change test, and an AIC value \> 10.
If so, we will include submodel 2 in our final model.

@@

A power analysis based on @kretzschmar_2019 indicated that with an alpha of α = 0.05, we will have \> 90% power to detect small effect sizes (β = 0.1) with a sample size ranging between N = 730 and N = 980 when the reliability of the measures is at least moderate (0.6 -- 0.7).
Therefore, we aim for a total sample size of N = 1,000.

@@

::: {#refs}
:::
